A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-18 paper/paper.tex
Minor

The relation $[[Interp I i A S]]$ satisfies the following inversion
principles.\scw{Put footnotes on lemma name, or on (1), (2), etc}
\begin{lemma}[Inversion of the logical relation]
  \label{lemma:interpinv}\leavevmode
  \begin{enumerate}
  \item If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Void\_inv}}
  \item If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}}
  \item If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}}
  \item If $[[Interp I i Pi A B S0]]$, then there exists $[[S]],[[F]]$ such that:\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}}
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
  \item If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Univ\_inv}}
  \end{enumerate}
\end{lemma}
\begin{proof}
  \scw{Why is this proof interesting and worth typesetting? Should
    it be in an appendix so that it doesn't interrupt the flow of 
    reading?}
  We only show the most involved function case; the rest follows a
  similar but simpler pattern. We start by
  inducting over the derivation of $[[Interp I i Pi A B S]]$. There
  are only two possible cases we need to consider.
  \begin{description}
  \item[\Rref{I-Pi}:] Immediate.
  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi A B S0]]$.
    We know that there exists some $[[A0]]$ and
    $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$ and $[[Interp I i Pi
    A0 B0 S0]]$. From the
    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
    By inverting $[[Pi A B => Pi A0 B0]]$, we derive $[[A => A0]]$ and
    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a} => B0 {a} ]]$ for all
    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
    $[[F]]$ satisfies the following properties:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
    These properties are exactly what's\scw{avoid contractions} required in the conclusion.
  \end{description}
\end{proof}

>>>>

The relation $[[Interp I i A S]]$ satisfies the following inversion
principles.\scw{Put footnotes on lemma name, or on (1), (2), etc}
\begin{lemma}[Inversion of the logical relation]
  \label{lemma:interpinv}\leavevmode
  \begin{enumerate}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Void\_inv}} If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}} If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
  \item If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}}
  \item If $[[Interp I i Pi A B S0]]$, then there exists $[[S]],[[F]]$ such that:\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}}
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
  \item If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Univ\_inv}}
  \end{enumerate}
\end{lemma}
\begin{proof}
  \scw{Why is this proof interesting and worth typesetting? Should
    it be in an appendix so that it doesn't interrupt the flow of 
    reading?}
  We only show the most involved function case; the rest follows a
  similar but simpler pattern. We start by
  inducting over the derivation of $[[Interp I i Pi A B S]]$. There
  are only two possible cases we need to consider.
  \begin{description}
  \item[\Rref{I-Pi}:] Immediate.
  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi A B S0]]$.
    We know that there exists some $[[A0]]$ and
    $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$ and $[[Interp I i Pi
    A0 B0 S0]]$. From the
    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
    By inverting $[[Pi A B => Pi A0 B0]]$, we derive $[[A => A0]]$ and
    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a} => B0 {a} ]]$ for all
    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
    $[[F]]$ satisfies the following properties:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
    These properties are exactly what's\scw{avoid contractions} required in the conclusion.
  \end{description}
\end{proof}
